Staged Abstract Interpreters

### 2019-09-06 paper/30_stageconc.tex
revision

In @close@, we first create a current-stage function @f@ that takes a
next-stage value and store, and returns staged values. Then we use
@emit_complied_clo@ to denote @f@ to a next-stage Scala function, represented
by the type @Rep[Value]@. Inside of @f@, we can access the evaluator via @ev@.
However, @ev@ produces a monadic value of type @AnsM@, which can only exist at
the current stage. To connect the current-stage monadic values and future-stage
grounded values, the evaluator @ev@ is invoked on the body expression @e@ to be
specialized w.r.t. @e@, and then we \textit{collapse} the monadic value of
@AnsM@ to grounded values by providing the prepared environment and store.

For function applications @ap_clo@, we have two \textit{next-stage} values @fun@
and @arg@. But what we can do on the values of @Rep@ type is limited. In fact, @fun@ does
not have an intensional application operation we can use directly. We only know
that it is a staged value produced by @close@, and represents a
next-stage Scala function. Therefore we can just generate a next-stage function
application, also in Scala. Since @ap_clo@ produces a monadic value of type
@Ans@, we can still use the monadic operation @get_store@ to obtain the latest
store object. Then @emit_ap_clo@ is used to generate a current-stage
representation of the future-stage application result, which is a pair of values
and stores @ValSt@. After getting that result @vs@, we use @put_store@ to
update the store and reify the result into the current-stage monadic value.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[ValSt]
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ  <- get_store
    vs <- lift[ValSt](emit_ap_clo(fun, arg, σ))
    _  <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

Compared with the unstaged version, we can observe a shift of using the
evaluation function @ev@ from application-time to value-representation-time: in
the unstaged interpreter, invocations of @ev@ happens in @ap_clo@; while in the
staged interpreter, we eagerly apply @ev@ when denoting lambda terms to our
next-stage value domain.

@@

In the previous section, we merely show the types of @emit_compiled_clo@ and
@emit_ap_clo@ without concrete implementations. In this section, we briefly
discuss the IR node generated by them and sketch code generation.
We define the IR nodes @IRCompiledClo@ and @IRApClo@ using @case class@
extending from @Def[T]@. @Def[T]@ is a built-in type from the LMS framework,
representing next-stage value definitions of type @T@.
\begin{lstlisting}
  case class IRCompiledClo(f: Rep[(ValSt) => ValSt], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Def[ValSt]
\end{lstlisting}

The IR nodes are manipulated by the LMS passes and finally generated to
next-stage Scala code. To generate code, LMS provides an @emitNode@ method for
programmers to control what is generated for each kind of nodes. Here we match
@IRCompiledClo@ and @IRApClo@ and generate definitions via @emitValDef@ for
them.  For compiled closures, we put the function @f@ and its accompanying
syntactic term and environment to its next-stage value representation
@CompiledClo@. For applications, we have the knowledge indicating that @fun@ is
a @CompiledClo@ at the stage, and it has a callable field @f@. Hence, we
generate a next-stage function application @f(arg, σ)@. The invocations of
@quote@s are used to generate code for the other next-stage values.
\begin{lstlisting}
override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
  case IRCompiledClo(f, λ, ρ) => emitValDef(sym, s"CompiledClo($\textdollar${quote(f)}, $\textdollar${quote(λ)}, $\textdollar${quote(ρ)})")
  case IRApClo(rator, rand, σ) => emitValDef(sym, s"$\textdollar$rator.f($\textdollar${quote(rand)}, $\textdollar${quote(σ)})")
  ...
}
\end{lstlisting}

>>>>

In @close@, we first create a current-stage function @f@ that takes a
next-stage value and store, and returns staged values. Then we use
@emit_complied_clo@ to denote @f@ to a next-stage Scala function, represented
by the type @Rep[Value]@. Inside of function @f@, we can access the evaluator via @ev@.
However, @ev@ produces a monadic value of type @AnsM@, which can only exist at
the current stage. To connect current-stage monadic values and future-stage
grounded values, the evaluator @ev@ is supplied with not only the body
expression @e@ to be specialized, but also the prepared environment and store.
By doing so, we \textit{collapse} the monadic value of @AnsM@ to grounded
values within staging; therefore in the future stage, there is no monads.

For function applications @ap_clo@, we have two \textit{next-stage} values @fun@
and @arg@. But what we can do with these values of type @Rep@ is limited. 
In fact, @fun@ does not have an intensional application operation we can use
directly. We only know that it is a staged value produced by @close@, and
it represents a next-stage Scala function. Therefore, we can only generate a
next-stage function application, also in Scala. @ap_clo@ produces a monadic
value of type @Ans@, so that we can still use the monadic operation @get_store@
to obtain the latest store. Then @emit_ap_clo@ takes the function and argument
values, and the latest store to generate a current-stage representation of the
future-stage application result, which is of type @Rep[ValSt]@, a pair of
values and stores.  After obtaining the result @vs@, we use @put_store@ to
update the store and reify the result into the current-stage monadic value.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[ValSt]
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ  <- get_store
    vs <- lift[ValSt](emit_ap_clo(fun, arg, σ))
    _  <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

Compared with the unstaged version, we can observe a shift of where we use the
evaluation function @ev@: in the unstaged interpreter, the invocation of @ev@
happens in @ap_clo@, i.e., at the application time; while in the staged
interpreter, we eagerly apply @ev@ when denoting the lambda terms to our
next-stage value domains, i.e., at the value-representation time.

@@

In the previous section, we showed the types of @emit_compiled_clo@ and
@emit_ap_clo@ without concrete implementations. In this section, we briefly
discuss the IR nodes generated by them and sketch the code generation.
We define the IR nodes @IRCompiledClo@ and @IRApClo@ using @case class@
extending from @Def[T]@. @Def[T]@ is a built-in type in the LMS framework,
representing next-stage value definitions of type @T@.
\begin{lstlisting}
  case class IRCompiledClo(f: Rep[(ValSt) => ValSt], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Def[ValSt]
\end{lstlisting}

The IR nodes are manipulated by the LMS passes and finally generated as
next-stage Scala code. To generate code, LMS provides an @emitNode@ method for
programmers to control over what code is generated for each kind of nodes. Here,
we match @IRCompiledClo@ and @IRApClo@, and generate definitions via
@emitValDef@ for them.  For compiled closures, we put the function @f@ and its
accompanying syntactic term and environment to their next-stage value
representation @CompiledClo@. For applications, we know the fact that @fun@ is
an instance of @CompiledClo@ at the next stage, and it has a callable field
@f@. Hence, we generate a next-stage function application @f(arg, σ)@.  The
invocations of @quote@ are used to generate proper names for the other
next-stage values.
\begin{lstlisting}
override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
  case IRCompiledClo(f, λ, ρ) => emitValDef(sym, s"CompiledClo($\textdollar${quote(f)}, $\textdollar${quote(λ)}, $\textdollar${quote(ρ)})")
  case IRApClo(rator, rand, σ) => emitValDef(sym, s"$\textdollar$rator.f($\textdollar${quote(rand)}, $\textdollar${quote(σ)})")
  ...
}
\end{lstlisting}

### 2019-09-06 paper/40_absinterp.tex
revision

After seeing the unstaged and staged concrete interpreter, now we turn our
focus on abstract interpreters under the same framework. We first present a
lattice representation with binding-time abstraction, as well as simple
abstract domains such as power set lattice.
Then we show the abstract components, namely abstract values, abstract
environments, and abstract stores.
The abstract interpreter we construct in this
section is similar to \citet{DBLP:journals/pacmpl/DaraisLNH17}'s. For
simplicity, it is also context/path/flow-insensitive by our choice. In Section
\ref{cfa}, we will instantiate it to a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

@@

Abstract interpreters operate on abstract domains, such abstract domains
usually can be formulated by complete lattices.
A complete lattice over set $S$ is a tuple of $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\top$ is the top element, $\bot$ is the
bottom element, $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines a type
class of stage polymorphic lattices. Similar to @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to
annotate the binding-times, where @R[_]@ wraps datatype @S@ and @Boolean@ in
these operations.

@@

In this section, we simply let the binding-time type @R[T] = T@ (i.e., @NoRep@)
to instantiate an unstaged lattice type class. An example of lattice we used in
the rest of the paper is the power set abstract domain (shown in Figure
\ref{fig:powerset}): the ordering is subset relation, the join is the union
operation, and the meet is the intersect operation. The bottom element of a
power set is empty set, and we don't have a top representation for power set.
Other lattices used in the paper, such as products and maps, can be
implemented similarly or lifted from existing lattices element-wise or
point-wise. Non-relational abstract domains such intervals can also be
implemented in a stage polymorphic way.

@@

We follow the \textit{abstracting definitional interpreters} (ADI) approach
\cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our concrete interpreter to
an abstract interpreter.
% \cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12}

@@

We first widen the concrete values to @AbsValue@: numbers are lifted to a
singleton object @IntTop@ representing the set of all numbers; closures remain
the same. Then the type @Value@ is a set of @AbsValue@ (we abuse the type names
a bit, as the generic interface already defined the abstract type @Value@).
The address space is constrained to be finite to ensure the computability of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After defining the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@, respectively.
\begin{lstlisting}
  trait AbstractComponents extends Semantics {
    sealed trait AbsValue
    case object IntTop extends AbsValue
    case class CloV(lam: Lam, env: Env) extends AbsValue
    type Value = Set[AbsValue]; case class Addr(x: String)
    ...
  }
\end{lstlisting}

It is important that the abstract store now maps addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
values occurred at run-time.  This can be justified by the approximation nature
of abstract interpretation and nondeterminism in the abstract interpreter.
First, when analyzing a conditional expression, we may not have enough
information on which branch will be taken, thus a sound treatment is to
nondeterministically explore both branches, i.e., compute the join of two path.
Second, when applying a function, for instance, @f(a)@, the possible lambda
terms of @f@ is computed approximately. As a result, we may retrieve a set of
closures when dereferencing the address of @f@ and need to apply all of them.

To prevent the abstract interpreter from falling into non-termination when
computing the fixed-point, the ADI approach uses a co-inductive mechanism
consisting of two caches that remember the input and output of functions.
Here, we first provide the necessary type definitions, and later we will describe
the algorithm in detail. A @Cache@ is a mapping from configurations @Config@ to
sets of value-store pairs, where the configuration is a triple of current
expression being evaluated, environment and store. Intuitively, a cache memoizes
the corresponding values and stores for a given program configuration.
\begin{lstlisting}
  type Config = (Expr, Env, Store); type Cache = Map[Config, Set[(Value, Store)]]
\end{lstlisting}

\paragraph{Monads for Abstract Interpretation} Compared with the concrete
interpreter that uses reader and state effects, the abstract interpreter
further introduces a nondeterminism effect and another reader and state effect
for the two caches. The nondeterminism effect is represented by the
@Set[M[_], A]@ monad transformer, where @M@ is the inner monad being
transformed, and @A@ is the type of elements in the set. We use a @ReaderT@ for
one cache that will not be changed during one fixed-point iteration, and use a
@StateT@ for another cache that will be constantly updated during the analysis.
\citet{DBLP:journals/pacmpl/DaraisLNH17} discuss different permutation of the
monad stack for abstract interpretation. The following @AnsM@ type shows the
monad stack (we use \hl{light gray} to highlight what have been changed from
the monad stack of concrete interpretation):
\begin{lstlisting}[escapechar=!]
trait AbstractSemantics extends AbstractComponents {
  type R[T] = T
  type AnsM[T] = ReaderT[StateT[!\hl{SetT}![!\hl{ReaderT}![!\hl{StateT}![IdM, !\hl{Cache}!, ?], !\hl{Cache}!, ?], ?], Store, ?], Env, T]
  ...
}
\end{lstlisting}

Similar to the concrete scenario, we sketch the implementation of @flatMap@ for
@SetT@ transformer and omit the rest. The value @run@ encapsulated by the monad
is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and @A@ is the element
type of the set. The @flatMap@ method takes a function that transforms each
element of type @A@ from @M[Set[A]]@ into a @SetT[M, B]@ value, and then folds
all the transformed values into a single monadic value of type @Set[M, B]@.
\begin{lstlisting}
  case class SetT[M[_]: Monad, A](run: M[Set[A]]) {
    def flatMap[B](f: A => SetT[M, B]): SetT[M, B] =
      SetT(Monad[M].flatMap(run) { (s: Set[A]) =>
        s.foldLeft(SetT.empty[M,B])((acc, a) => acc ++ f(a)).run
      }); ... }
\end{lstlisting}

Such monad stack scheme leads to a different type of grounded result. In the concrete
setting, we only have a reader monad (for environment) and a state monad (for
store), and the reader effect is not persistent to the final result. Together
with the value produced by the interpreter, the final result type is a pair of
@Value@ and @Store@. In the abstract setting, we have a @SetT@ inside of the
environment monad and store monad, which becomes the container type of pairs
of values and stores, i.e., @Set[(Value, Store)]@. Also note that the reader and
state monad for caches inhibit internally of the nondeterminism monad, therefore
the cache type of state monad becomes adjacent to the set type.
\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

\paragraph{Primitive Operations} The primitive operations are changed according to
the new monad stack scheme and value domains. A notable change is that the store
update operation joins the new value with existing values, for the same address.
As we mentioned before, we may have to explore both branches for conditionals:
in @br0@ method, we combine the results using @mplus@ from @MonadPlus@, which
requires the value domains are join-semilattices (in our case @Set@s and
@Map@s).
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}

The value representation of lambda terms is still a defunctionalized closure
@CloV@, but we lift it to a singleton set domain.
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = Set(CloV(λ, ρ))
\end{lstlisting}

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is that now the first argument @rator@ is a set of closures (remember
that @Value@ is an alias of @Set[AbsValue]@), so we can not directly
destruct it by pattern matching. Instead, we use an auxiliary function @lift_nd@
that takes a set and lifts the elements into the monad stack. Then we can still
implement @ap_clo@ in monadic style, where the closures come from the @SetT@ and
thus the nondeterminism can be naturally handled. The light gray code shows what
is added from the concrete @ap_clo@.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](rator)}!
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

@@

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination, which is also called a
\textit{co-inductive} caching algorithm or \textit{truncated depth-first
evaluation} \cite{Rosendahl:AbsIntPL}. It has been used in other abstract
interpreter or fixed point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use a @in@ cache and a @out@ cache during
the depth-first evaluation, where the @in@ cache stores the result from last
iteration, and @out@ cache is used for updating in current iteration. In the next
iteration, the last @out@ cache will be served as @in@ cache, and an empty cache
is plugged-in to the @out@ slot. Once the @out@ cache does not contain anything
new than the @in@ cache, the fixed-point is reached.

In our monad stack, the @in@ and @out@ cache are modeled by a reader effect and
state effect, respectively. We first define several methods to manipulate
the two caches through the monad stack (the implementations are elided).
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

The co-inductive caching algorithm is implemented as a instrumentation over
@eval@ (Figure \ref{fig:coind_cache}), and it also closes the open recursion.
Initially, the two caches are both empty.
During the iteration, we first check whether @out@ contains the configuration
@cfg@, representing the current desired computation. If it does, then the result
is directly returned through the monads.
Otherwise, we first retrieve the results from @in@ (@⊥@ if not contained in
@in@), and join it into the @out@ cache.
Then we invoke @ev@ to compute the result for this iteration, where @ev@ takes
@fix(ev)@ as a self reference.
After the evaluation, the store @σ@ may have been changed. We update the
@out@ cache with the new result, i.e., a pair of value and store @(v, σ)@.
Similar to store update, cache update is also a point-wise join.

The iteration terminates when the resulting @out@ cache is equivalent to the
input @in@ cache, which indicates that there is no more fact can be discovered,
therefore the iteration should end and we have reached the least fixed point.

>>>>

After seeing the unstaged and staged concrete interpreter, now we turn our
focus to abstract interpreters under the same framework. We first present a
lattice representation with binding-time abstraction, as well as simple
abstract domains, such as power set lattice.  Then we show the abstract
components, namely abstract values, abstract environments, and abstract stores.
The abstract interpreter we construct in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s. For simplicity, it is
context/path/flow-insensitive by our choice. In Section \ref{cfa}, we will
instantiate it to a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} show how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our
unstaged or staged abstract interpreter.

@@

The abstract domains operated by abstract interpreters usually can be
formulated by complete lattices.
A complete lattice over set $S$ is a tuple $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\top$ is the top element, $\bot$ is the
bottom element, $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines a type
class of stage polymorphic lattices. Similar to @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to
annotate the binding-times, where @R[_]@ wraps the data type @S@ and @Boolean@
in these operations.

@@

In this section, we simply instantiate @SPLattice@ with the flat binding-time type
@R[T] = T@ (i.e., @NoRep@).
An example of lattices we used in the rest of the paper is the power set
abstract domain (shown in Figure \ref{fig:powerset}): the ordering is the subset
relation, the join is the union operation, and the meet is the intersect
operation. The bottom element of a power set is empty set, and we do not have a
representation of top element for power set.  Other lattices used in the paper, such as
products and maps, can be implemented similarly or by lifting the existing
lattices element-wise or point-wise. Non-relational numerical abstract domains
such intervals can also be implemented in a stage polymorphic way.

@@

In this section, we follow the \textit{abstracting definitional interpreters}
(ADI) approach \cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our concrete
interpreter to an abstract interpreter.

@@

We first widen the concrete values to @AbsValue@. There are two variants of
@AbsValue@: numbers are lifted to a singleton object @IntTop@, which represents
the set of all numbers; closures remain the same. Then the type @Value@ is
defined as a set of @AbsValue@.  The address space is constrained to be finite
to ensure the computability of analysis -- for a simple monovariant analysis,
we directly use variable names for addresses. After defining the @Value@ and
@Addr@ type, the types of environments and stores are automatically lifted to
abstract environments and abstract stores, i.e., @Map[Ident, Addr]@ and
@Map[Addr, Set[AbsValue]]@, respectively.  
\begin{lstlisting}
  trait AbstractComponents extends Semantics {
    sealed trait AbsValue
    case object IntTop extends AbsValue
    case class CloV(lam: Lam, env: Env) extends AbsValue
    type Value = Set[AbsValue]; case class Addr(x: String)
    ...
  }
\end{lstlisting}

It is important that the abstract stores now map addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
runtime values.  This can be justified by the approximation nature of abstract
interpretation and nondeterminism in the abstract interpreter.  First, when
analyzing a conditional expression, we may not have enough information to
decide which branch will be taken, thus a sound treatment is to explore the both
branches. Also, at some later point, we need to compute the join of two path.
Another nondeterminism comes from function applications, for instance, @f(a)@,
the possible callees ($\lambda$ terms) of @f@ is computed approximately.  As a
result, we may retrieve a set of closures when dereferencing the address of @f@
and we need to apply all of them.

To ensure the abstract interpreter terminates on all programs when computing
the fixed-point, the ADI approach uses a co-inductive mechanism consisting of
two caches that remember the input and output of the @eval@ function.
Here, we first provide the necessary type definitions, and describe
the algorithm later. A @Cache@ is a mapping from configurations @Config@ to
sets of value-store pairs, where the configuration is a triple of the current
expression being evaluated, environment and store. Intuitively, a cache memoizes
the result values and stores for a given program configuration.
\begin{lstlisting}
  type Config = (Expr, Env, Store); type Cache = Map[Config, Set[(Value, Store)]]
\end{lstlisting}

\paragraph{Monads for Abstract Interpretation}
Compared with the concrete interpreter that uses reader and state effects, the
abstract interpreter further introduces the nondeterminism effect and another
reader and state effect for the two caches. The nondeterminism effect is
represented by the @Set[M[_], A]@ monad transformer, where @M@ is the inner
monad type being transformed, and @A@ is the type of elements in the set. We use a
@ReaderT@ for one cache that is not changed during one fixed-point
iteration, and use a @StateT@ for another cache that will be constantly updated
during the analysis.  \citet{DBLP:journals/pacmpl/DaraisLNH17} discuss
different permutations of the monad stack for abstract interpretation. The
following @AnsM@ type shows the monad stack for abstract interpretation (we use
\hl{light gray} to highlight what have been changed from the monad stack of
concrete interpretation):
\begin{lstlisting}[escapechar=!]
trait AbstractSemantics extends AbstractComponents {
  type R[T] = T
  type AnsM[T] = ReaderT[StateT[!\hl{SetT}![!\hl{ReaderT}![!\hl{StateT}![IdM, !\hl{Cache}!, ?], !\hl{Cache}!, ?], ?], Store, ?], Env, T]
  ...
}
\end{lstlisting}

Similar to the concrete scenario, we sketch the @flatMap@ implementation of the
@SetT@ transformer and omit the rest. The field @run@ encapsulated by the monad
is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and @A@ is the element
type of the set. The @flatMap@ method takes a function that transforms each
element of type @A@ into a @SetT[M, B]@ value, and then folds
all the transformed values into a single monadic value of type @Set[M, B]@.
\begin{lstlisting}
  case class SetT[M[_]: Monad, A](run: M[Set[A]]) {
    def flatMap[B](f: A => SetT[M, B]): SetT[M, B] =
      SetT(Monad[M].flatMap(run) { (s: Set[A]) =>
        s.foldLeft(SetT.empty[M, B])((acc, a) => acc ++ f(a)).run
      }); ... }
\end{lstlisting}

Such monad stack scheme leads to a different type of grounded result. In the concrete
setting, we only have a reader monad (for environments) and a state monad (for
stores). The reader effect is not persistent to the final result. Together
with the value produced by the interpreter, the final result type is a pair of
@Value@ and @Store@. However, under abstract interpretation, we have a @SetT@
inside of the environment monad and store monad. The @Set@ type becomes the
container type of the pairs of values and stores, i.e., @Set[(Value, Store)]@.
Also, note that the reader and state monad for the caches inhibits internally
of the nondeterminism monad; therefore, the final result type is paired
with a cache type from the state monad.
\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

\paragraph{Primitive Operations} The primitive operations are changed according to
the new monad stack scheme and value domains. One of the notable changes is
that the store update operator merges the new values with existing values.
As we mentioned before, sometimes we have to explore the both branches for
conditionals: in method @br0@, we combine the results using the @mplus@
operation from @MonadPlus@, which requires the value domains are
join-semilattices (in our case, @Set@ and @Map@).
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}

The value representation of lambda terms is still a defunctionalized closure
@CloV@, but we lift it to a singleton set.
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = Set(CloV(λ, ρ))
\end{lstlisting}

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is that now the first argument @funs@ is a set of closures (remember
that type @Value@ is an alias of @Set[AbsValue]@), so we can not directly
destruct it to syntactic @Lam@ by pattern matching. Instead, we use an
auxiliary function @lift_nd@ that takes a set and lifts the elements in the set
into the monad stack. Then we can straightforwardly implement the @ap_clo@ in
monadic style, where the closures come from the @SetT@ and thus the
nondeterminism can be naturally handled. The light gray code shows what is
added from the concrete
@ap_clo@.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(funs: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](funs)}!
    α <- alloc(x)
    _ <- set_store(α → funs)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

@@

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination.
The caching algorithm is also called \textit{co-inductive} caching or
\textit{truncated depth-first evaluation} \cite{Rosendahl:AbsIntPL}. It has
been used in other abstract interpreter or fixed point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use a @in@ cache and a @out@ cache during
the depth-first evaluation. The @in@ cache stores the result from the last
iteration, and @out@ cache is used for updating in the current iteration. In the next
iteration, the last @out@ cache will be served as @in@ cache, and an empty cache
is plugged-in to the @out@ slot. Once the @out@ cache does not contain anything
new than the @in@ cache, the fixed-point is reached.

In our monad stack, the @in@ and @out@ caches are modeled by reader effects and
state effects, respectively. We first define several methods to manipulate
the two caches through the monad stack (the implementations are elided).
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

The co-inductive caching algorithm is implemented as an instrumentation over the
@eval@ function (Figure \ref{fig:coind_cache}), and it also closes the open recursion.
The instrumentation works as follows. Initially, the two caches are both empty.
During the iteration, we first check whether @out@ contains the configuration
@cfg@, representing the current desired computation. If it does, then the result
is directly returned through the monad stack.
Otherwise, we first retrieve the result from @in@ (@⊥@ if @in@ not contains
@cfg@), and update the result into the @out@ cache in the fashion of join.
Then, we invoke @ev@ to compute the result for this iteration, where @ev@ takes
@fix(ev)@ as a self reference.
After the evaluation, the store @σ@ may have been changed. Therefore, 
we get the latest store and construct the result @(v, $\sigma$)@; the
result will be used to update the out cache.

The iteration terminates when the resulting @out@ cache is equivalent to the
input @in@ cache, which indicates that there is no more fact can be discovered.
Therefore, the iteration should end, and we have reached the least fixed point.
